Best Practices for Alchemical Free Energy Calculations: v\versionnumber

### 2020-01-27 paper/manuscript.tex
typo fix

Alchemical free energy calculations compute free energy differences associated with transfer processes, such as the binding of a small molecule to a receptor, the transfer of a small molecule from an aqueous to apolar phase~\cite{zwanzig1954hightemperature}, or effects of protein side chain mutations on e.g. binding affinities.. 

@@

Alchemical calculations can be used in a variety of scenarios, such as computing the free energy of a conformational change for a molecule with a high barrier to interconversion (Fig.~\ref{fig:fig_what_is_alchemy} (\textbf{A})); 

@@

computing partition ($\log P$) or distribution ($\log D$) coefficients between environments (Fig.~\ref{fig:fig_what_is_alchemy} (\textbf{B}))~\cite{rustenburg2016measuring, bosisio2016blinded} or compartments or membranes (Fig.~\ref{fig:fig_what_is_alchemy} (\textbf{C}))~\cite{corey2019insights}. 
Furthermore, alchemical calculations are frequently used to estimate changes in free energies upon modifying a ligand or protein: a protein residue can be alchemically mutated to probe the impact on binding affinity (Fig.~\ref{fig:fig_what_is_alchemy} (\textbf{D}))\cite{hauser2018predicting,aldeghi2018accurate} or changes in protein thermostability~\cite{seeliger2010protein,gapsys2016insights,gapsys2016accurate,aldeghi2019accurate}; the entire ligand can be alchemical transferred from protein to solvent in an absolute binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} (\textbf{E}))~\cite{mobley2007predicting,aldeghi2015accurate,aldeghi2017predictions}; and small alchemical modifications can be made between chemically related ligands to estimate differences in binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} (\textbf{F}))~\cite{wang2015accurate,mey2016blinded,gapsys2020large}.

@@

Once an alchemical simulation as been performed, the data must be analyzed to estimate the free energy for the transformation of interest.
Early work used simple (but suboptimal) estimators for this: free energy perturbation (FEP) using the Zwanzig relation~\cite{zwanzig1954hightemperature} or thermodynamic integration (TI), for which the theory dates back decades but with the first computational applications emerging in the 1980's and 90's~\cite{kirkwood1935statistical, jorgensen1985monte, kollman1993free, wong1986dynamics, merz1989free}. %

@@

\todo[inline, color={yellow!20}]{JDC: The key pivotal moment was the Shirts paper on extremely precise free energy calculations, which demonstrated that a combination of technologies can produce robust estimates with useful precision.}
Important studies continued during the 2000's often leading to improved implementation of the methods in common simulation software~\cite{vanderspoel2005gromacs, mermelstein2018fast, wang2015accurate, hedges2019biosimspace}. 

@@

As the field of molecular simulation moves towards millisecond timescales, accurate alchemical calculations on even more challenging problems will become reasonable to perform. In the meantime, today's users may find it hard to get started with these complex calculations whilst also keeping up with the fast pace of change. This Best Practice guide aims to serve as a go-to reference for current recommendations and tips for users of all experience.  

@@

    \caption{\textbf{Illustration of common types of free energies differences that can be calculated using alchemical free energy methods.} \textbf{A}: Change in free energy due to a conformational change of the molecule. \textbf{B}: Partition coefficient such as $\log P$ or $\log D$ depend on a change in free energy between different phases; here, as an example the partition coefficient between methanol and water is shown. \textbf{C}: Free energy difference associated with the insertion of a molecule into a membrane. \textbf{D}: Effect of mutations of protein or host residues on free energies of binding. \textbf{E}: Absolute free energies of binding of a small molecule to a host (e.g. protein), \textbf{F}: Relative free energy of binding of one molecule with respect to another, here toluene and benzyl alcohol.

@@

This guide is meant to serve as a comprehensive starting point for new practitioners for free energy calculations.  It assumes moderate experience with molecular simulation packages and concepts, however.  It is also intended to serve as a reference guide with a convenient checklist (Section~\ref{sec:checklist}) for more seasoned practitioners, to help ensure their simulation meet standard alchemical simulation and analysis practices. It can also serve as a set of best practices to ensure simulation robustness and reproducibility which reviewers may wish to consider as they evaluate papers.
This guide aims to answer the following questions:
    \item Is my problem suitable for an alchemical free energy calculation? 
    \item How do I choose the right free energy protocol and run it? 
    \item How do I analyze my simulations accurately? 
    \item What software tools are available to perform alchemical free energy calculations? 
We assume the reader possesses a basic familiarity with the principles of molecular mechanics, molecular dynamics simulations, statistical mechanics, and the biophysics of protein-ligand association. If you feel unfamiliar with some of these concepts please good starting points can be found in these references~\cite{braun2019best, grossfield2018best, klimovich2015guidelines, shirts2012best}. Oftentimes, docking calculations are required to generated initial binding pose for a small molecule for an alchemical calculation, where binding poses are not known, thus some basic familiarty on how to run docking calculations is also expected~\cite{grinter2014challenges}. 
As some of the theoretical background can seem daunting, we provide an essentials guide to the theory behind alchemical free energy calculations in Section~\ref{sec:theory}.
In the remainder of this paper, we will cover topics that are key to the preparation~(\ref{sec:prerequisites}), choice and use of correct protocols~(\ref{sec:simulation_protocol_choice}), and finally the best practices that should be used for the analysis of alchemical calculations~(\ref{sec:data_analysis}). 
Particular focus will be given to aspects of the molecular simulations which are unique to alchemical calculations---these include the calculation of transfer free energies (hydration free energies, partition coefficients, etc.), and binding free energies (absolute and relative).
While we try to address as many methods and practices as possible, the field of free energy calculations is broad, and there are many advanced topics that are left to future best practices documents focusing on specific issues. 
Below, we provide a non-exhaustive list of topics we have not addressed with some references to provide starting points on these more advanced topics:
\item Covalent inhibition.
\item Free energies of mutation of protein side chains
\item nonspecific binding or multiple binding sites.
\item Complex association stoichiometry (beyond simple 1:1 association).
\item Endpoint free energy methods such as MM-PBSA~\cite{genheden2015mm} and LIE~\cite{gutierrez-de-teran2012linear}
\item Free energy methods that extract the ligand using geometric order parameters and potential of mean force methods~\cite{heinzelmann2017attachpullrelease}
\item Forcefield dependence for protein, ligand, ions, co-solvents, and co-factors. Different studies have looked at the influence of force fields and it is assumed the user has made an adequate choice for the system under study.~\cite{loeffler2018reproducibility, vassetti2019assessment, lopes2015current} 
For convenience we have also compiled a list of common acronyms, common symbols used, and some essential terminology.
\begin{tcolorbox}[title=Acronyms, colback=blue!10!white]
    \textit{CPU} -- Central Processing Unit\\
    \textit{BAR} -- Bennett Acceptance Ratio\\
    \textit{FEP} -- Free Energy Perturbation\\
    \textit{GPCR} -- G-Protein Coupled Receptor\\
    \textit{GPU} -- Graphics Processing Unit\\
    \textit{MBAR} -- Multistate Bennett Acceptance Ratio\\
    \textit{MCSS} -- Maximum Common Substructure\\
    \textit{MD} -- Molecular Dynamics\\
    \textit{RMSE} -- Root Mean Square Error\\
    \textit{MUE} -- Mean unsigned error\\
    \textit{SAR} -- Structure-Activity Relationships\\
    \textit{TI} -- Thermodynamic Integration
\end{tcolorbox}
%\label{sec:abbreviations}
\begin{tcolorbox}[title=List of Symbols, colback=green!10!white]
$U$ -- potential energy\\
$u$ -- reduced potential energy \\
$\Delta G$ -- free energy \\
$\Delta f$-- reduced free energy \\
$k_B$-- Boltzmann constant \\
$T$ -- temperature
$k_b$ -- rate of binding \\
$K_D$ -- dissociation constant \\
$\lambda$ -- parameter for regulation of the alchemcial free energy path \\
$g$ -- statistical inefficiency
\end{tcolorbox}
\begin{tcolorbox}[title=Common terminology, colback=yellow!10!white]
    ligand - small molecular binding to a larger host molecule such as a protein (recpetor).\\ 
    lead optimization - process of designing a drug candidate from an initial lead compound, considering its affinity and various pharmacokinetic parameters.\\
    thermodynamic cycle - sequence of thermodynamic processes (like changes in temperature, volume or pressure) in which the initial and final state are equal.\\
\end{tcolorbox}
% Theory basics    %
\section{Statistical mechanics can tell us why alchemical free energy calculations work}
\label{sec:theory}
Why would you want to run an alchemical free energy calculation and why do they work? Suppose you want to compute the binding affinity, or free energy of binding, of a ligand $L$ to a protein $P$, given by:
P+L\leftrightharpoons PL
The equilibrium constant  ($K_b^{\circ}$) is given by the ratio of concentrations of product $[PL]$ and reactants $[P]$, $[L]$:
 K_b^{\circ} = c^{\circ}\frac{[PL]}{[L][P]}.
In order to define a reference state, the standard state is usually defined as $1 \mathrm{mol}/\mathrm{L}$ and $c^\circ$ is the standard state concentration. Thus, the Gibbs free energy of binding $\Delta G_{\mathrm{bind}}$ is given by:
    \Delta G_{\mathrm{bind}} = -k_BT\ln K_b^{\circ},
where $k_B$ is the Boltzmann constant and $T$ the temperature of the system. 
One way to estimate the equilibrium constant is by computing the probability of finding the molecular system in the bound state ($PL$) or the unbound state ($P+L$), making use of partition function ratios, and assuming that the contribution of water compressibility at 1 atm has a negligible effect on the free energy of binding~\cite{shirts2012best, gilson1997statisticalthermodynamic} i.e.

\begin{equation}
    \Delta G_{\mathrm{bind}} \cong \Delta F= -k_BT\ln\frac{\int_\mathrm{bound} \exp(-\beta U(\vec{q}))d\vec{q}}{\int_\mathrm{unbound} \exp(-\beta U(\vec{q}))d\vec{q}},
\end{equation}
where $\beta = \frac{1}{k_BT}$ and $F$ is the Helmholtz free energy. 
Or more generally in a simplified form the free energy difference ($\Delta f$) between the bound and unbound states is:
\begin{equation}
\Delta f \equiv f(\mathrm{bound}) - f(\mathrm{unbound}) = -\ln\frac{Z(\mathrm{unbound})}{Z({\mathrm{bound}})},
\end{equation}
where $Z = \int_\mathrm{bound} \exp(-\beta U(\vec{q}))d\vec{q}$ represents the configurational integral. 
Unfortuantely it is not possible to evaluate the integral required for the configurational integral exactly, but can be approximated using molecular dynamics simulations. 
So, how can we compute $\Delta G_{\mathrm{bind}}$ from a molecular dynamics simulation? Running a direct simulation of the binding events is too computationally costly, meaning we cannot easily estimate the configurational integrals of the bound and unbound state from the simulation, as typically binding events happen at a $\mu$s scale, and simulations reach around 1 $\mu$s/day. For a reliable estimate of a single free energy of binding many days of simulations would be required, making these type of calculations uninteresting at a pharamaceutical timescale. Instead, we can run alchemical free energy simulations, by making use of the fact that free energy is a state function. We can write down a thermodynamic cycle as seen in ~\ref{fig:fig_binding_thermodynamic_cycle}, for a relative free energy simulation. Based on this cycle we can compute the relative free energy of binding between two compounds $A$ and $B$, by going along the arrow of $\Delta G_{\mathrm{bound}}$, where we introduce a second compound that is similar to the first one. Now $\Delta G_{\mathrm{bound}}$ can be expressed as a ratio of partition functions of toluene (A) and benzyl alcohol (B). 
The defining characteristic of alchemical free energy calculations is the use of a series of alchemically-modified potential functions $U(x; \lambda)$ in which an alchemical parameter $\lambda$ modulates interactions in a manner that cannot occur in real chemical systems.
One or more simulations are used to collect data from a multitude of alchemical states to compute a free energy difference between a chemical state ($\lambda_0$) and another chemical or alchemical reference state ($\lambda_1$),
\begin{eqnarray}
\Delta f &\equiv& f(\lambda_1) - f(\lambda_0) = - \ln \frac{Z(\lambda_1)}{Z(\lambda_0)} , \label{equation:dimensionless-free-energy-difference}
\end{eqnarray}
where the dimensionless free energy $f(\lambda) \equiv \beta F(\lambda)$ is given in terms of partition functions $Z(\lambda)$,
\begin{eqnarray}
Z(\lambda) &=& \int_{\Gamma} d\vec{q} \, \exp({-u(\vec{q}; \lambda)}) ,
\label{equation:partition-function-definition}
\end{eqnarray}
, where $\Gamma$ signifies the phase space volume over which to integrate. 
The \emph{reduced potential} $u(\vec{q}; \lambda)$~\cite{shirts2008statisticallya} is generally given by a trace over thermodynamic parameters with their conjugate dynamical variables,
\begin{eqnarray}
u(\vec{q};\lambda) &\equiv& \beta \left[ U(\vec{q};\lambda) + p \, V(\vec{q}) + \sum_{i=1}^N \mu_i \, N_i(\vec{q}) + \cdots \right] . \label{equation:reduced-potential}
\end{eqnarray}
Here, the collection of thermodynamic and alchemical parameters $\theta \equiv \{\beta, \lambda, p, \mu, \ldots\}$ defines a \emph{thermodynamic state}.
The physical transformation of interest---such as a protein-ligand association process, or transfer free energy among phases---may involve a thermodynamic cycle that requires several free energy differences $\Delta f$ to be computed in order to produce an estimate of the overall desired quantity, see for example the thermodynamic cycle for an absolute free energy calculation as seen in~\ref{fig:fig_absolute_thermodynamic_cycle}.

@@

%%%%%%%%%%%%
% Step 0   %
%%%%%%%%%%%%
\section{What can be expected from alchemical simulations?}
\label{sec:step0}
\todo[inline, color={green!20}]{ASJSM: @everyone please review this section}
When starting an alchemical free energy project you need to determine if an alchemical free energy calculation is the appropriate strategy by initially counting the cost of your
project: Can you even hope to tackle the problem with available resources and, if successful, will it
be worth it?
\subsection{How accurate are alchemical free energy calculations?}
\label{subsec:expectation}
Current alchemical free energy calculations seem to achieve, in favorable cases, RMS errors around 1-2 kcal/mol depending on force field, system, and a variety of other factors such as simulation time, sampling method, and whether the calculations employed are absolute or relative. A small selection of benchmark datasets and case studies can be found in section~\ref{sec:benchmark} at the end of this document.
However, the domain of applicability is a significant concern~\cite{sherborne2016collaborating, cournia2017relative}, especially for relative calculations, which typically require a bound structure of a closely
related ligand as a starting point. Additional factors such as slow protein or ligand rearrangements, uncertainties in ligand binding mode, or charged ligands can make these calculations far less reliable and more of a research effort.
It is worth noting that the accuracy of free energy calculations is highly variable across different protein targets, and likely across different ligand chemotypes as well. For instance, FEP+ with OPLS3 achieves an RMSE of 0.62 kcal/mol for a set of 21 compounds binding to JNK1 kinase, yet only an RMSE of 1.05 kcal/mol for a set of 34 compounds binding to P38$\alpha$ kinase~\cite{harder2016opls3}. Furthermore, perturbations for the same chemotype in different pockets of the BACE enzyme gave varied errors~\cite{keranen2017acylguanidine}
\todo[inline,color={green!20}]{JDC: It's important to be clear on what error is being reported: $\Delta G$ after shifting by a constant to minimize the RMSE, unshifted $\Delta G$, $\Delta \Delta G$ of computed edges, or $\Delta \Delta G$ of all edges. Note that computed edge $\Delta \Delta G$ RMSE significantly underestimates the error of all edge RMSE: \url{https://github.com/jchodera/jacs-dataset-analysis} ASJSM: Can we incorporate your analysis in this section as an example? Maybe this is something that needs to be addressed properly in the analysis section and only referenced here?}

>>>>

Alchemical free energy calculations compute free energy differences associated with transfer processes, such as the binding of a small molecule to a receptor, the transfer of a small molecule from an aqueous to apolar phase~\cite{zwanzig1954hightemperature}, or effects of protein side chain mutations on e.g. binding affinities. 

@@

Alchemical calculations can be used in a variety of scenarios, such as: 
\item computing the free energy of a conformational change for a molecule with a high barrier to interconversion (Fig.~\ref{fig:fig_what_is_alchemy} (\textbf{A}));

@@

\item computing partition ($\log P$) or distribution ($\log D$) coefficients between environments (Fig.~\ref{fig:fig_what_is_alchemy} (\textbf{B}))~\cite{rustenburg2016measuring, bosisio2016blinded} 
\item or between compartments or membranes (Fig.~\ref{fig:fig_what_is_alchemy} (\textbf{C}))~\cite{corey2019insights}. 

Furthermore, alchemical calculations are frequently used to estimate changes in free energies upon modifying a ligand or protein: 
\item a protein residue can be alchemically mutated to probe the impact on binding affinity (Fig.~\ref{fig:fig_what_is_alchemy} (\textbf{D}))\cite{hauser2018predicting,aldeghi2018accurate} or changes in protein thermostability~\cite{seeliger2010protein,gapsys2016insights,gapsys2016accurate,aldeghi2019accurate}; 
\item the entire ligand can be alchemical transferred from protein to solvent in an absolute binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} (\textbf{E}))~\cite{mobley2007predicting,aldeghi2015accurate,aldeghi2017predictions}; 
\item and small alchemical modifications can be made between chemically related ligands to estimate differences in binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} (\textbf{F}))~\cite{wang2015accurate,mey2016blinded,gapsys2020large}.

@@

Once an alchemical simulation procedure is performed (which usually involves more than one simulation), the data must be analyzed to estimate the free energy for the transformation of interest.
Early work used simple but statistically suboptimal estimators for this: free energy perturbation (FEP) using the Zwanzig relation~\cite{zwanzig1954hightemperature} or thermodynamic integration (TI), for which the theory dates back decades but with the first computational applications emerging in the 1980's and 90's~\cite{kirkwood1935statistical, jorgensen1985monte, kollman1993free, wong1986dynamics, merz1989free}. %

@@

\todo[inline, color={yellow!20}]{JDC: The key pivotal moment was the Shirts paper on extremely precise free energy calculations, which demonstrated that a combination of technologies can produce robust estimates with useful precision. MRS: aw, shucks! added.}
Important studies continued during the 2000's often leading to improved implementation of the methods in common simulation software~\cite{shirts2003extremely,shirts2005solvation,vanderspoel2005gromacs, mermelstein2018fast, wang2015accurate, hedges2019biosimspace}. 

@@

As the field of molecular simulation moves towards millisecond timescales, accurate alchemical calculations on even more challenging problems will become reasonable to perform. In the meantime, today's users may find it hard to get started with these complex calculations whilst also keeping up with the fast pace of change. This Best Practice guide provides current recommendations and tips for users of all experience.  

@@

    \caption{\textbf{Illustration of common types of free energies differences that can be calculated using alchemical free energy methods.} \textbf{A}: Change in free energy due to a conformational change of the molecule across a high barrier. \textbf{B}: Partition coefficient such as $\log P$ or $\log D$ depend on a change in free energy between different phases; here, as an example the partition coefficient between methanol and water is shown. \textbf{C}: Free energy difference associated with the insertion of a molecule into a membrane. \textbf{D}: Effect of mutations of protein or host residues on free energies of binding. \textbf{E}: Absolute free energies of binding of a small molecule to a host (e.g. protein), \textbf{F}: Relative free energy of binding of one molecule with respect to another, here toluene and benzyl alcohol.

@@

This guide is meant to serve as a comprehensive starting point for new practitioners of free energy calculations.  However, it assumes moderate experience with molecular simulation packages and concepts.  It is also intended to serve as a reference guide with a convenient checklist (Section~\ref{sec:checklist}) for more seasoned practitioners, to help ensure their calculations meet standard alchemical simulation and analysis practices. It can also serve as a set of best practices to ensure simulation robustness and reproducibility which reviewers may wish to consider as they evaluate papers. Importantly, it is not, a review of all free energy calculation methods at the cutting edge of current research.

This guide aims to answer the following questions:
\begin{itemize}
    \item Is my problem suitable for an alchemical free energy calculation? 
    \item How do I choose the right free energy protocol and run it? 
    \item How do I analyze my simulations accurately? 
    \item What software tools are available to perform alchemical free energy calculations? 
\end{itemize}
%
We assume the reader possesses a basic familiarity with the principles of molecular mechanics, molecular dynamics simulations, statistical mechanics, and the biophysics of protein-ligand association. If you feel unfamiliar with some of these concepts, good starting points can be found in these references~\cite{braun2019best, grossfield2018best, klimovich2015guidelines, shirts2012best}.

Other background information may be needed depending on the project. For example, often, if binding poses are not known, docking calculations can be used to generate an initial small molecule binding pose to start alchemical calculations, thus some basic familiarity on how to perform docking is also necessary for such projects~\cite{grinter2014challenges}. 
%
As some of the theoretical background can seem daunting, in this best practices description, we do provide a guide to the essential theory behind alchemical free energy calculations in Section~\ref{sec:theory}.

In the remainder of this paper, we will cover topics that are key to the preparation~(\ref{sec:prerequisites}), choice and use of correct protocols~(\ref{sec:simulation_protocol_choice}), and finally the best practices that should be used for the analysis of alchemical calculations~(\ref{sec:data_analysis}). 
Particular focus will be given to aspects of the molecular simulations which are unique to alchemical calculations---these include the calculation of transfer free energies (hydration free energies, partition coefficients, etc.), and binding free energies (absolute and relative).
%

While we try to address as many methods and practices as possible, the field of free energy calculations is broad, and there are many advanced topics that are left to future best practices documents focusing on specific issues. 
Below, we provide a non-exhaustive list of topics we have not addressed with some references to provide starting points on these more advanced topics:
\begin{itemize}
\item Covalent inhibition.
\item Free energies of mutation of protein side chains
\item nonspecific binding or multiple binding sites.
\item Complex association stoichiometry (beyond simple 1:1 association).
\item Approximate endpoint free energy methods such as MM-PBSA~\cite{genheden2015mm} and LIE~\cite{gutierrez-de-teran2012linear}
\item Free energy methods that extract the ligand using geometric order parameters and potential of mean force methods~\cite{heinzelmann2017attachpullrelease}
\item Forcefield dependence for protein, ligand, ions, co-solvents, and co-factors. A number of different studies have looked at the influence of force fields and it is assumed the user has made an adequate choice for the system under study.~\cite{loeffler2018reproducibility, vassetti2019assessment, lopes2015current} 
\end{itemize}
%
For convenience we have also compiled a list of common acronyms, common symbols used, and some essential terminology.
\begin{tcolorbox}[title=Acronyms, colback=blue!10!white]
    \textit{CPU} -- Central Processing Unit\\
    \textit{BAR} -- Bennett Acceptance Ratio\\
    \textit{FEP} -- Free Energy Perturbation\\
    \textit{GPCR} -- G-Protein Coupled Receptor\\
    \textit{GPU} -- Graphics Processing Unit\\
    \textit{MBAR} -- Multistate Bennett Acceptance Ratio\\
    \textit{MCSS} -- Maximum Common Substructure\\
    \textit{MD} -- Molecular Dynamics\\
    \textit{RMSE} -- Root Mean Square Error\\
    \textit{MUE} -- Mean unsigned error\\
    \textit{SAR} -- Structure-Activity Relationships\\
    \textit{TI} -- Thermodynamic Integration
\end{tcolorbox}
%\label{sec:abbreviations}
\begin{tcolorbox}[title=List of Symbols, colback=green!10!white]
$U$ -- potential energy\\
$u$ -- reduced potential energy \\
$\Delta G$ -- free energy \\
$\Delta f$-- reduced free energy \\
$K_b$ -- binding or association constant \\
$k_B$-- Boltzmann constant \\
$T$ -- temperature \\
$\beta$ -- (k_BT)^{-1}
$k_b$ -- rate of binding \\
$K_D$ -- dissociation constant \\
$\lambda$ -- parameter for progress of the alchemical free energy path \\
$g$ -- statistical inefficiency
\end{tcolorbox}
\begin{tcolorbox}[title=Common terminology, colback=yellow!10!white]
    \emph{ligand} - small molecule binding to a larger host molecule such as a protein (receptor).\\ 
    \emph{lead optimization} - process of designing a drug candidate from an initial lead compound, usually involving optimizing its affinity, possibly selectivity, and various pharmacokinetic parameters.\\
    \emph{thermodynamic cycle} - sequence of thermodynamic processes (like changes in temperature, volume or pressure) in which the initial and final state are equal.\\
\end{tcolorbox}
%
%%%%%%%%%%%%%%%%%%%%
% Theory basics    %
%%%%%%%%%%%%%%%%%%%%
\section{Statistical mechanics demonstrates why alchemical free energy calculations work}
\label{sec:theory}
Why would you want to run an alchemical free energy calculation and why do they work?
In this section, we use the example of relative free energy calculation to sketch the theory of alchemical calculations and illustrate their utility.
The emphasis here is placed on bridging theoretical foundations and intuition.
A rigorous derivation of the standard (absolute) free energy of binding using the principles of statistical mechanics can be found in Gilson's classic work~\cite{gilson1997statisticalthermodynamic}.

\subsection{Simulating binding events of receptor-drug systems can be computationally expensive}

Suppose you want to compute the binding affinity, or free energy of binding, of a ligand $L$ to a receptor $R$, given by:
\begin{equation}
R+L \leftrightharpoons RL
\end{equation}
The binding constant  ($K_b^{\circ}$) is given by the law of mass action as the ratio of concentrations of product $[RL]$ and reactants $[R]$, $[L]$:
\begin{equation}
 K_b^{\circ} = c^{\circ}\frac{[RL]}{[L][R]}.
\end{equation}
The standard state concentration $c^\circ$ depends on the reference state, but it is usually set to $1\,\mathrm{mol}/\mathrm{L}$ assuming a constant pressure of $1\,\mathrm{atm}$ (see also Section~\ref{sec:standardstate-restraints}).
Thus, the Gibbs free energy of binding $\Delta G_{\mathrm{bind}}$ is given by:
\begin{equation}
    \Delta G_{\mathrm{bind},L} = -k_BT\ln K_b^{\circ},
\end{equation}
where $k_B$ is the Boltzmann constant and $T$ the temperature of the system.

\paragraph{The free energy of binding can be expressed as a ratio of partition functions}
A natural though generally very computational expensive way to estimate the equilibrium constant is by directly simulating several binding and unbinding events and computing the probability of finding the receptor-ligand system in the bound state, $P(RL)$, or the unbound state, $P(R+L)$.
Assuming the volume change upon binding to be negligible, which is often the case at $1~\mathrm{atm}$ due to the incompressibility of water, then the Gibbs free energy $\Delta G_{\mathrm{bind}, L}$ is approximately equal to the Helmholtz free energy $\Delta A_{\mathrm{bind}, L}$, and we can simulate the system in a box of volume $V$ to obtain~\cite{jong_determining_2011}
\begin{equation}\label{eq:binding-free-energy-from-bound-unbound-probability-ratio}
    \Delta G_{\mathrm{bind},L} \approx \Delta A_{\mathrm{bind}, L} = - k_B T  \left( \mathrm{ln} \frac{P(RL)}{P(R+L)} + \ln \left( c^\circ N_{\mathrm{Av}} V \right) \right) \, ,
\end{equation}
where $N_{\mathrm{Av}}$ is the Avogadro number, and the last term corrects for the simulated concentration being different than the standard concentration.
Let $\Gamma_{\mathrm{bound}}$ and $\Gamma_{\mathrm{unbound}}$ be the set of receptor-ligand conformations $\vec{q}$ that we consider bound and unbound respectively.
The probability of a conformation $\vec{q}$ is given by the Boltzmann probability density function
\begin{equation}
    P(\vec{q}) = \frac{\exp\left( \beta U(\vec{q}) \right)}{\int_{\Gamma} \exp\left( \beta U(\vec{q}) \right) \, d\vec{q}} \, ,
\end{equation}
where $\beta = (k_B T)^{-1}$ is the inverse temperature, $U(\vec{q})$ is the potential energy of conformation $\vec{q}$, and the integration is over the set of all possible conformations accessible in the simulation box volume $\Gamma$, with $\Gamma_{\mathrm{bound}},\Gamma_{\mathrm{unbound}} \subset \Gamma$.
If the simulation is long enough, we expect the fraction of conformations $\vec{q}$ found in the bound state to converge to
\begin{equation}
    P(RL) = \int_{\Gamma_{\mathrm{bound}}} P(\vec{q}) \, d\vec{q} = \frac{\int_{\Gamma_{\mathrm{bound}}} \exp\left( \beta U(\vec{q}) \right) \, d\vec{q}}{\int_{\Gamma} \exp\left( \beta U(\vec{q}) \right) \, d\vec{q}} \, .
\end{equation}
After similar considerations for $P(R+L)$, we find that the ratio of visited bound and unbound conformations, in the limit of long simulations, should converge to
\begin{equation}\label{eq:bound-unbound-probability-ratio}
    \frac{P(RL)}{P(R+L)} = \frac{\int_{\Gamma_{\mathrm{bound}}} \exp\left( -\beta U(\vec{q}) \right) \, d\vec{q}}{\int_{\Gamma_{\mathrm{unbound}}} \exp\left( -\beta U(\vec{q}) \right) \, d\vec{q}} = \frac{Z(RL)}{Z(R+L)} \, ,
\end{equation}
where we have defined the \textit{configurational integral} or \textit{partition function} $Z(\mathrm{state}) \equiv \int_{\Gamma_{\mathrm{state}}} \exp\left(-\beta U(\vec{q})\right) \, d\vec{q}$.

\paragraph{Simulating binding events is computationally expensive}
While simulating binding events has been used to estimate binding affinities~\cite{jong_determining_2011,pan2017quantitative} or to get insights into the binding pathways and kinetics of receptor-ligand systems~\cite{teo_adaptive_2016,votapka_seekr_2017,doerr_--fly_2014,plattner_protein_2015,dixon_predicting_2018}, the computational cost of these calculations is usually dominated by the rate of dissociation, which can be on the $\mu$s scale even for millimolar binders~\cite{pan2017quantitative} and reaches the ms or s scale for a typical drug~\cite{basavapathruni_conformational_2012,hyre_cooperative_2006}.
Depending on system size and simulation settings, common molecular dynamics software packages can reach a few hundreds of ns/day using currently available high-end GPUs~\cite{eastman_openmm_2017,kutzner_more_2019}, making these type of calculations uninteresting at pharmaceutical drug discovery timescales.
Other methods compute the free energy of binding by building potential of mean force profiles along a reaction coordinate~\cite{woo2005calculationa,velezvega_overcoming_2013,limongelli_funnel_2013,heinzelmann2017attachpullrelease}, but these methods require prior knowledge of a high-probability binding pathway, which is not easily available, especially in the prospective scenarios typical of the drug development process.

\subsection{Alchemical free energy calculations yield predictions that do not require direct simulation of binding events}

In many cases, the quantity of interest is the change in binding affinity when a compound $A$ is mutated to compound $B$ (e.g., by modifying one the drug scaffold's substituents), which, by using Eq.~\ref{eq:binding-free-energy-from-bound-unbound-probability-ratio} and \ref{eq:bound-unbound-probability-ratio} is given by
\begin{equation}\label{eq:delta-delta-G-physical}
\begin{split}
    \Delta \Delta G_{\mathrm{bind}, AB} &= \Delta G_{\mathrm{bind}, B} - \Delta G_{\mathrm{bind}, A} \\
    &\approx -k_BT \left( \ln \frac{Z(RB)}{Z(R+B)} - \ln \frac{Z(RA)}{Z(R+A)} \right) \, .
\end{split}
\end{equation}
Note that the terms involving the standard concentration cancel out when we assume that the volume is identical for $A$ and $B$.
Predictions of $\Delta \Delta G_{\mathrm{bind}, AB}$ with non-alchemical methods generally require long simulations of both ligands, possibly through different binding pathways.
Alchemical relative free energy calculations avoid the need to simulate binding and unbinding events by making use of the fact that the free energy is a state function and exploiting the thermodynamic cycle illustrated in Figure~\ref{fig:fig_binding_thermodynamic_cycle}.
This is apparent after rewriting Eq.~\ref{eq:delta-delta-G-physical} as
\begin{equation}\label{eq:delta-delta-G-alchemical}
\begin{split}
    \Delta \Delta G_{\mathrm{bind}, AB} &\approx -k_BT \left( \ln \frac{Z(RB)}{Z(RA)} - \ln \frac{Z(R+B)}{Z(R+A)} \right) \\
    &= -k_BT \left( \ln \frac{Z(RB)}{Z(RA)} - \ln \frac{Z(B)}{Z(A)} \right) \\
    &= \Delta G_{\mathrm{bound}} - \Delta G_{\mathrm{unbound}} \, ,
\end{split}
\end{equation}
where $\Delta G_{\mathrm{bound/unbound}}$ is the free energy of mutating $A$ to $B$ in the bound/unbound state.
Eq.~\ref{eq:delta-delta-G-alchemical} and Figure~\ref{fig:fig_binding_thermodynamic_cycle} tell us that the difference in free energy of binding between toluene ($A$) and benzyl alcohol ($B$) can be computed by running two independent calculations estimating the free energy cost of mutating $A$ into $B$ in the binding pocket ($\Delta G_{\mathrm{bound}}$) and in solvent ($\Delta G_{\mathrm{unbound}}$), saving us the need to simulate the physical binding process of the two compounds.
In particular, the second line of Eq.~\ref{eq:delta-delta-G-alchemical} is a consequence of $\Delta G_{\mathrm{unbound}}$ being independent of the presence of the receptor in the simulation box as the definition of the unbound state assumes receptor and ligand to be at a sufficient distance for them to have no energetic interactions.
Note that, when $A$ and $B$ have different number of atoms, Eq.~\ref{eq:delta-delta-G-alchemical} implies the presence of a factor having units of volume entering both logarithms, which require unitless arguments.
The factor's value is, however, inconsequential as it cancels out in practice.

@@

\paragraph{How are alchemical transformations performed in practice?}

In practice, the mutation of $A$ to $B$ in carried out by introducing one or more parameters $\vec{\lambda}$ controlling the potential energy function $U(\vec{q};\vec{\lambda})$ such that the potential of compounds $A$ and $B$ is recovered at two particular values $\vec{\lambda}_A$ and $\vec{\lambda}_B$.
Briefly, this is achieved by simulating a ``chimera'' molecule composed by enough atoms to represent both $A$ and $B$.
A subset of the energetic terms in $U(\vec{q};\vec{\lambda})$ is then modulated by $\vec{\lambda}$ so that at $\vec{\lambda}_A$, the atoms that form molecule $A$ are activated and those belonging exclusively to $B$ are non-interacting ``dummy atoms'', while the opposite occur at $\lambda_B$ (see Section~\ref{sec:relative-fe-protocol} for details).
